Boosted Mirror Descent, Frank-Wolfe, and Herding

### 2013-10-03 icml/paper.tex
small notational changes and other updates; switched to icml2014 style files

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2013 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}
\usepackage{import, icml2013}
\input{latex-defs.tex}

@@

\icmlauthor{Jacob Steinhardt}{jsteinhardt@cs.stanford.edu}
\icmladdress{Stanford University,
             353 Serra Street, Stanford, CA 94305 USA}
\icmlauthor{Jonathan Huggins}{jhuggins@mit.edu}
\icmladdress{Massachusetts Institute of Technology,
             43 Vassar Street, Cambridge, MA 02139 USA}

@@

\NA{JHH: I'm thinking that from the beginning we should let $u$ be defined over a different space than $\theta$ and assume a linear map $u \mapsto \eta(u) \in \Theta$. Then write $\langle \theta, \eta(u) \rangle$ so the setup will align with the herding case, where $\eta(u) = \bE_{u}[\phi(x)]$. }

Let $U$ and $\Theta$ be convex subsets of two Hilbert spaces $\sH_{U}$ and $\sH_{\Theta}$ and 
let $i : U \to \Theta$ be a linear map from 
$U$ to $\Theta$ that preserves the inner product. Consider the following loss function 
$L : U \to \bR$:
\begin{equation}
L(u) = h(u) + \max_{\theta \in \Theta} \left\{\langle \theta, i(u) \rangle - R(\theta) \right\}.
\end{equation}
We can think of $h$ as a \emph{primal regularizer} and $R$ as a \emph{dual regularizer}.

Examples:
\begin{itemize}
\item Let $U$ be the probability simplex over a space $\sX$, let $\phi : \sX \to \bR^d$ be a collection 
      of statistics, and let $\Theta$ be the unit $L^1$ ball in $\bR^d$. Let $h(u) = 0$, 
      $\langle \theta, i(u) \rangle = \bE_u[\theta^T\phi(x)]$, and 
      $R(\theta) = \langle \theta, i(u_0) \rangle$. Then $L(u)$ is the 
      \emph{maximum mean discrepancy} between $u$ and $u_0$ relative to $\phi$.
\item Let $h(u) = 0$ and $R(\theta) = S^*(i^{-1}(\theta))$ where $S^*$ is the 
      Fenchel conjugate of any strongly convex function $S : U \to \bR$. Then $L(u) = S(u)$ 
      as long as $i^{-1}(\Theta) = \sH_{U}$.
\item Let $U = \Theta = \bR^d$. Let $h(u) = \|u\|_1$, $R(u) = \langle \theta, u_0 \rangle + \frac{1}{2} \|\theta\|_2^2$. 
      Then $L(u) = \|u\|_1 + \frac{1}{2} \|u-u_0\|_2^2$. 
\end{itemize}

@@

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\begin{equation}
L(u,\theta) = h(u) + \langle \theta, i(u) \rangle - R(\theta).
\end{equation}
We will assume throughout that $h$ and $R$ are both convex functions, 
and furthermore that $\arg\max_{u} h(u) + \langle \theta, i(u) \rangle$ can be efficiently 
computed for all values of $\theta$.

Mirror descent, together with boosting, yields an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm (called Algorithm 1).
For notational convenience, for a sequence of weights $\alpha_1,\alpha_2,\ldots$ 
let $\hat{u}_t = \frac{\sum_{s=1}^t \alpha_su_s}{\sum_{s=1}^t \alpha_s}$ and let 
$\hat{\theta}_t = \frac{\sum_{s=1}^t \alpha_s\theta_s}{\sum_{s=1}^t \alpha_s}$.
\begin{enumerate}
\item $u_1 \in \arg\min_u h(u)$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, i(u_t) \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \hat{\theta}_t, i(u) \rangle$
%\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
As long as $h$ is strongly convex, we obtain the 
bound (see Corollary~\ref{cor:method-1}):
\begin{equation}
\sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T).
\end{equation}
In other words, $\hat{u}_T$ is close to being a global minimum of $L$.
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm (called Algorithm 2):
\begin{enumerate}
\item $\theta_1 \in \arg\min_{\theta} R(\theta)$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, i(u) \rangle$
%\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, i(\hat{u}_t) \rangle - R(\theta)$
\end{enumerate}
We then obtain the following bound 
when $R$ is strongly convex (see Corollary~\ref{cor:method-2}):
\[ \sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T). \]
Algorithm 1 and Algorithm 2 are closely related; indeed, they are dual to each other 
(performing Algorithm 1 on $L(u,\theta)$ is the same as performing Algorithm 2 on 
$-L(\theta,u)$).

For convenience, we will abuse notation and use $\langle \theta, u \rangle$ in place of 
$\langle \theta, i(u) \rangle$ when the map $i$ is clear from context.

@@

Bach et al. have observed that herding appears to be approximately building 
samples from the maximum entropy distribution for a given collection of moment 
constraints. In this section we will make this connection more explicit 
as well as draw other relationships between herding and maximum entropy.

Suppose that instead of setting $h(u) = 0$ we set $h(u)$ to a small multiple of 
the negative entropy: $h(u) = \eta \bE_{u}[\log u(x)]$. We will still use 
$R(\theta) = \frac{1}{2}\|\theta\|_2^2$. In this case, Algorithm 2 gives the updates
\begin{itemize}
\item $u_t(x) \propto \exp\left(-\frac{1}{\eta}\theta_t^T\phi(x)\right)$
\item $\theta_{t+1} = \bE_{\hat{u}_t}[\phi(x)]$.
\end{itemize}
Using Corollary~\ref{cor:method-2} gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T u_t, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2r^2\log(T+1)}{T} \\
 &= \inf_{u^*} \left\{-\eta H(u^*) + \frac{1}{2}\|\bE_{u^*}[\phi(x)]-\bar{\phi}\|_2^2\right\} \\ &\phantom{=} + \frac{2r^2\log(T+1)}{T}.
\end{align*}
In particular, let $u^*$ be the maximum entropy distribution satisfying $\bE_{u^*}[\phi(x)] = \bar{\phi}$, 
and let $H^*$ be the corresponding entropy. Then we have 
\[ -\eta H(\hat{u}) + \|\bE_{\hat{u}}[\phi(x)]-\bar{\phi}\|_2^2 \leq -\eta H^* + \frac{2r^2\log(T+1)}{T}. \]
Since $\|\cdot\|_2^2 \geq 0$, this implies that 
\begin{equation}
H(\hat{u}) \geq H^* - \frac{2r^2\log(T+1)}{\eta T},
\end{equation}
 and so asymptotically $\hat{u}$ has at least 
as much entropy as $u^*$.

Now, suppose that our state space has some finite size $S$, so 
that no distribution has entropy greater than $\log(S)$. Then 
we conclude that
\begin{equation}
\|\bE_{\hat{u}}[\phi(x)]-\bar{\phi}\|_2^2 \leq \eta[\log(S)-H^*] + \frac{2r^2\log(T+1)}{T}.
\end{equation}

@@

%% Let us now consider the entropic regularizer $h(u) = \tau\bE_{u}[\log u(x)]$ with 
%% \[ R(\theta) = \bE_{u}[\theta^T\bar{\phi}] + \left\{ \begin{array}{llc} \infty & : & \|\theta\|_1 > 1 \\ 0 & : & \|\theta\|_1 \leq 1 \end{array} \right.\]
%% If we use Method 1, we then end up with 
%% \[ u_{t}(x) \propto \exp\left\{-\frac{1}{\alpha}\overline{\theta}_{t}^T\phi(x)\right\}, \]
%% where $\overline{\theta}_{t} = \frac{1}{t} \sum_{s \le t} \theta_{s}$. Hence, $h(u) + \bE_{\mu}[\theta^T\phi(x)] - R(\theta)$ is no longer maximized at the boundary (when $\tau > 0$). We can instead consider the following sequence of distributions:
%% \begin{enumerate}
%% \item $\theta_t \in \argmax_{\|\theta\|_1 \leq 1} \theta^T (\bE_{u_t}[\phi(x)]-\bar{\phi})$
%% \item $u_{t+1}(x) = Z(\overline{\theta}_{t}, \tau)^{-1} \exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\}$
%% \end{enumerate}
%% Here $Z(\overline{\theta}_{t}, \tau) = \int\exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\} \dee x$. 
%% 
%% The entropic regularizer is not strongly convex but a more general analysis involving 
%% Bregman divergence can nevertheless yield a convergence bound in this case. We thus 
%% get something that looks like herding but actually gives us an exponential family 
%% type distribution. In particular, we have a Gibbs distribution with temperature $\tau$, and 
%% in the limit as the temperature goes to zero, we recover herding, since 
%% \[ \lim_{\tau \to 0} u_{t+1}(x) = \delta_{x^{*}_{t+1}}, \]
%% where $x^{*}_{t+1} = \argmin_{x} \overline{\theta}_{t}^T\phi(x)$. 
%% This fact is reminiscent of Welling's original derivation of herding, in which 
%% he took the zero temperature limit of the gradient descent algorithm for the optimization problem 
%% \[ \argmin_{u} h(u) \quad \text{s.t.}~\bE_{u}[\phi(x)] - \bar\phi = 0. \]
%% This optimization is quite similar to the entropic optimization considered here, which 
%% we can write as 
%% \begin{align*}
%% \argmin_{u}& \argmax_{|\theta\|_1 \leq 1} h(u) + \theta^{T}(\bE_{u}[\phi(x)] - \bar \phi)  \\
%% &= \argmin_{u} h(u) + \|\bE_{u}[\phi(x)] - \bar\phi\|_{2}. 
%% \end{align*}
%% In the entropic formulation, the expectation constraint $\bE_{u}[\phi(x)] - \bar\phi = 0$ becomes
%% soft, though the two optimizations are equivalent in the zero-temperature limit. 
%% 
%% We can view the entropic regularization
%% with non-zero temperature as an alternative approximation to herding. If at each step we
%% retain a sample $x_{t} \sim u_{t}$, then the sequence $(x_{t})_{t}$ converges to the 
%% herding sample path as $\tau \to 0$. So $(x_{t})_{t}$ with non-zero temperature approximate
%% herding samples. \NA{JHH: I'm not sure how useful this all is, other than to make contact with 
%% the original paper, as I think the Gibbs distribution was exactly what Max was trying to 
%% avoid by taking the zero-temperature limit.}

>>>>

\documentclass{article}

\usepackage{hyperref}

\usepackage{icml2014}

\input{latex-defs.tex}

@@

\icmlauthor{Jonathan Huggins \NA{add footnote on authorship}}{jhuggins@mit.edu}
\icmladdress{MIT CSAIL,
             77 Massachusetts Avenue, Cambridge, MA 02139 USA}
\icmlauthor{Jacob Steinhardt}{jsteinhardt@cs.stanford.edu}
\icmladdress{Stanford University,
             353 Serra Street, Stanford, CA 94305 USA}

@@

Let $U$ and $\Theta$ be convex subsets of two Banach spaces $\sB_{U}$ and $\sB_{\Theta}$ and 
let $\imath : U \to \Theta$ be a linear map from 
$U$ to $\Theta$ that preserves the inner product. Consider the following loss function 
$L : U \to \bR$:
\begin{equation}
L(u) = h(u) + \max_{\theta \in \Theta} \left\{\langle \theta, \imath(u) \rangle - R(\theta) \right\}.
\end{equation}
We can think of $h$ as a \emph{primal regularizer} and $R$ as a \emph{dual regularizer}.

Examples:
\begin{itemize}
\item Let $U$ be the probability simplex over a space $\sX$, let $\phi : \sX \to \bR^d$ be a collection 
      of statistics, and let $\Theta$ be the unit $L^1$ ball in $\bR^d$. Let $h(u) = 0$, 
      $\langle \theta, \imath(u) \rangle = \bE_u[\theta^T\phi(x)]$, and 
      $R(\theta) = \langle \theta, \imath(u_0) \rangle$. Then $L(u)$ is the 
      \emph{maximum mean discrepancy} between $u$ and $u_0$ relative to $\phi$.
\item Let $h(u) = 0$ and $R(\theta) = S^*(i^{-1}(\theta))$ where $S^*$ is the 
      Fenchel conjugate of any strongly convex function $S : U \to \bR$. Then $L(u) = S(u)$ 
      as long as $i^{-1}(\Theta) = \sB_{U}$.
\item Let $U = \Theta = \bR^d$. Let $h(u) = \|u\|_1$, $R(u) = \langle \theta, u_0 \rangle + \frac{1}{2} \|\theta\|_2^2$. 
      Then $L(u) = \|u\|_1 + \frac{1}{2} \|u-u_0\|_2^2$. 
\end{itemize}

@@

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\begin{equation}
L(u,\theta) = h(u) + \langle \theta, \imath(u) \rangle - R(\theta).
\end{equation}
We will assume throughout that $h$ and $R$ are both convex functions, 
and furthermore that $\arg\max_{u} h(u) + \langle \theta, \imath(u) \rangle$ can be efficiently 
computed for all values of $\theta$.

Mirror descent, together with boosting, yields an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm (called Algorithm 1).
For notational convenience, for a sequence of weights $\alpha_1,\alpha_2,\ldots$ 
let $\hat{u}_t = \frac{\sum_{s=1}^t \alpha_su_s}{\sum_{s=1}^t \alpha_s}$ and let 
$\hat{\theta}_t = \frac{\sum_{s=1}^t \alpha_s\theta_s}{\sum_{s=1}^t \alpha_s}$.
\begin{enumerate}
\item $u_1 \in \arg\min_u h(u)$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, \imath(u_t) \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \hat{\theta}_t, \imath(u) \rangle$
%\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
As long as $h$ is strongly convex, we obtain the 
bound (see Corollary~\ref{cor:method-1}):
\begin{equation}
\sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T).
\end{equation}
In other words, $\hat{u}_T$ is close to being a global minimum of $L$.
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm (called Algorithm 2):
\begin{enumerate}
\item $\theta_1 \in \arg\min_{\theta} R(\theta)$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, \imath(u) \rangle$
%\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \imath(\hat{u}_t) \rangle - R(\theta)$
\end{enumerate}
We then obtain the following bound 
when $R$ is strongly convex (see Corollary~\ref{cor:method-2}):
\[ \sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T). \]
Algorithm 1 and Algorithm 2 are closely related; indeed, they are dual to each other 
(performing Algorithm 1 on $L(u,\theta)$ is the same as performing Algorithm 2 on 
$-L(\theta,u)$).

For convenience, we will abuse notation and use $\langle \theta, u \rangle$ in place of 
$\langle \theta, \imath(u) \rangle$ when the map $i$ is clear from context.

@@

Bach et al.~have observed that herding appears to be approximately building 
samples from the maximum entropy distribution for a given collection of moment 
constraints. In this section we will make this connection more explicit 
as well as draw other relationships between herding and maximum entropy.

Suppose that instead of setting $h(u) = 0$ we set $h(u)$ to a small multiple of 
the negative entropy: $h(u) = \eta \bE_{u}[\log u(x)]$. We will still use 
$R(\theta) = \frac{1}{2}\|\theta\|_2^2$. In this case, Algorithm 2 gives the updates
\begin{itemize}
\item $u_t(x) \propto \exp\left(-\frac{1}{\eta}\theta_t^T\phi(x)\right)$
\item $\theta_{t+1} = \bE_{\hat{u}_t}[\phi(x)] - \bar\phi$.
\end{itemize}
Corollary~\ref{cor:method-2} implies that for all $u^{*}$,
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T u_t, \theta\right) &\leq \sup_{\theta} L(u^*, \theta) + \frac{2r^2\log(T+1)}{T} \\
 &= -\eta H(u^*) + \frac{1}{2}\|\bE_{u^*}[\phi(x)]-\bar{\phi}\|_2^2 \\ &\phantom{=} + \frac{8r^2}{T}.
\end{align*}
In particular, let $u^*$ be the maximum entropy distribution satisfying $\bE_{u^*}[\phi(x)] = \bar{\phi}$, 
and let $H^*$ be the corresponding entropy. Then we have 
\[ -\eta H(\hat{u}) + \|\bE_{\hat{u}}[\phi(x)]-\bar{\phi}\|_2^2 \leq -\eta H^* + \frac{8r^2}{T}. \]
Since $\|\cdot\|_2^2 \geq 0$, this implies that 
\begin{equation}
H(\hat{u}) \geq H^* - \frac{8r^2}{\eta T},
\end{equation}
and so asymptotically $\hat{u}$ has at least 
as much entropy as $u^*$.

Now, suppose that our state space has some finite size $S$, so 
that no distribution has entropy greater than $\log(S)$. Then 
we conclude that
\begin{equation}
\|\bE_{\hat{u}}[\phi(x)]-\bar{\phi}\|_2^2 \leq \eta[\log(S)-H^*] + \frac{8r^2}{T}.
\end{equation}

@@

%\subsection{Old stuff}
%
% Let us now consider the entropic regularizer $h(u) = \tau\bE_{u}[\log u(x)]$ with 
% \[ R(\theta) = \bE_{u}[\theta^T\bar{\phi}] + \left\{ \begin{array}{llc} \infty & : & \|\theta\|_1 > 1 \\ 0 & : & \|\theta\|_1 \leq 1 \end{array} \right.\]
% If we use Method 1, we then end up with 
% \[ u_{t}(x) \propto \exp\left\{-\frac{1}{\alpha}\overline{\theta}_{t}^T\phi(x)\right\}, \]
% where $\overline{\theta}_{t} = \frac{1}{t} \sum_{s \le t} \theta_{s}$. Hence, $h(u) + \bE_{\mu}[\theta^T\phi(x)] - R(\theta)$ is no longer maximized at the boundary (when $\tau > 0$). We can instead consider the following sequence of distributions:
% \begin{enumerate}
% \item $\theta_t \in \argmax_{\|\theta\|_1 \leq 1} \theta^T (\bE_{u_t}[\phi(x)]-\bar{\phi})$
% \item $u_{t+1}(x) = Z(\overline{\theta}_{t}, \tau)^{-1} \exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\}$
% \end{enumerate}
% Here $Z(\overline{\theta}_{t}, \tau) = \int\exp\left\{-\frac{1}{\tau}\overline{\theta}_{t}^T\phi(x)\right\} \dee x$. 
% 
% The entropic regularizer is not strongly convex but a more general analysis involving 
% Bregman divergence can nevertheless yield a convergence bound in this case. We thus 
% get something that looks like herding but actually gives us an exponential family 
% type distribution. In particular, we have a Gibbs distribution with temperature $\tau$, and 
% in the limit as the temperature goes to zero, we recover herding, since 
% \[ \lim_{\tau \to 0} u_{t+1}(x) = \delta_{x^{*}_{t+1}}, \]
% where $x^{*}_{t+1} = \argmin_{x} \overline{\theta}_{t}^T\phi(x)$. 
% This fact is reminiscent of Welling's original derivation of herding, in which 
% he took the zero temperature limit of the gradient descent algorithm for the optimization problem 
% \[ \argmin_{u} h(u) \quad \text{s.t.}~\bE_{u}[\phi(x)] - \bar\phi = 0. \]
% This optimization is quite similar to the entropic optimization considered here, which 
% we can write as 
% \begin{align*}
% \argmin_{u}& \argmax_{|\theta\|_1 \leq 1} h(u) + \theta^{T}(\bE_{u}[\phi(x)] - \bar \phi)  \\
% &= \argmin_{u} h(u) + \|\bE_{u}[\phi(x)] - \bar\phi\|_{2}. 
% \end{align*}
% In the entropic formulation, the expectation constraint $\bE_{u}[\phi(x)] - \bar\phi = 0$ becomes
% soft, though the two optimizations are equivalent in the zero-temperature limit. 
% 
% We can view the entropic regularization
% with non-zero temperature as an alternative approximation to herding. If at each step we
% retain a sample $x_{t} \sim u_{t}$, then the sequence $(x_{t})_{t}$ converges to the 
% herding sample path as $\tau \to 0$. So $(x_{t})_{t}$ with non-zero temperature approximate
% herding samples. \NA{JHH: I'm not sure how useful this all is, other than to make contact with 
% the original paper, as I think the Gibbs distribution was exactly what Max was trying to 
% avoid by taking the zero-temperature limit.}
